---
title: "Clustering through the epigraph and hypograph indices"
output: 
 rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Clustering through the epigraph and hypograph indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 78
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, warning = FALSE, error = FALSE,
  fig.cap = "",
  fig.path = "includes/ehymet-02-"
)
```

```{r setup}
set.seed(42)

library(ehymet)
```

# The EHyClus function

One of the main utilities of the epigraph and the hypograph index is to be
able to use classical clustering methods with them. is to be able to use
classical clustering methods with them. In the package, the function that
allows us to perform this process is `EHyClus`.

The EHyClus function is designed for clustering multivariate datasets that
contain curves, where the input data can be one-dimensional (n x p) or
multi-dimensional (n x p x q). The function constructs a dataset using the
different indices discussed on the first vignette such as the epigraph,
hypograph, and their modified versions, and then applies various clustering
algorithms to this dataset.

The function can automatically determine the best combinations of variables
for clustering (any number can be selected but the default is one) or use
specified combinations provided by the user.

It supports multiple clustering methods: **hierarchical clustering**
("hierarch"), **k-means** ("kmeans"), **kernel k-means** ("kkmeans"), and
**spectral clustering** ("spc"). Also, it allows customization of hierarchical
clustering methods, distance metrics, and kernels using parameters like
`l_dist_hierarch`.

For smoothing, it uses B-splines with a specified or automatically selected
number of basis functions.

To check the quality of the results, if true labels are provided, the function
can validate the clustering results and compute performance metrics such as
purity, F-measure, and Rand Index (RI). Also, it records the time taken for
each clustering method, in case you want to measure it in case of wanting to
know the tradeoff between results and execution time

## Parameters

All the parameters and its functionality can be found on the function
documentation, but arguably the most important ones are:

-   **curves**: The dataset containing the curves to be clustered.
-   **vars_combinations**: Determines the combinations of variables to use.
    Can be an integer for automatic selection or a list of specific
    combinations.
-   **clustering_methods**: A vector specifying which clustering methods to
    apply.
-   **k**: Number of basis functions for B-splines.
-   **bs**: Type of penalized smoothing basis to use.
-   **indices**: Indices to generate for clustering.

## Example of usage and validation

In this subsection, we are going to display how to obtain the results of the
`EHyClus` function and validate them (if `true_labels` are present). First, we
are going to generate multidimensional data using `sim_model_ex2`:

```{r}
n <- 50
curves <- sim_model_ex2(n = n, i_sim = 4)
```

And, as we are generating $n = 50$ curves per group and there are two groups,
the true labels are:

```{r}
true_labels <- c(rep(1, n), rep(2, n))
```

We can run the algorithm with or without the true labels. For the combinations
of variables, we will use "d2dtaMEI" with "d2dtaMHI".

```{r}
res <- EHyClus(curves, vars_combinations = list(c("d2dtaMEI", "d2dtaMHI")))
res_with_labels <- EHyClus(curves,
  vars_combinations = list(c("d2dtaMEI", "d2dtaMHI")),
  true_labels = true_labels
)
```

For the result without labels, it can be seen that we only have the generated
clusters:

```{r}
names(res)
```

Whereas the one we generated with labels has both `cluster` and `metrics`:

```{r}
names(res_with_labels)
```

Let's explore both components of the latter. We're going to start with
`cluster`.

```{r, echo=FALSE}
data.tree::FromListSimple(res_with_labels$cluster, nodeName = "cluster")
```

The first thing that stands out is that the combination of variables used,
i.e. the one chosen as the best, is "dtaMEIdtaMHI". This can be seen in the
prefix of the different elements. We can also see the different parameters
used. For example, for the kmeans it is easy to see that it has been performed
with both the Euclidean and the Mahalanobis distance.

Looking in particular at some of the elements, we see that it contains the
following:

```{r}
str(res_with_labels$cluster$hierarch$hierarch_ward.D2_euclidean_d2dtaMEId2dtaMHI)
```

On the one hand we have `cluster` which is the vector that assigns each of the
curves to a cluster. Then we have `valid`, which is the validation data. And
finally `time`, which is the time that this particular method has taken to be
executed. Let's take a look at `valid`:

```{r}
head(res_with_labels$cluster$hierarch$hierarch_ward.D2_euclidean_d2dtaMEId2dtaMHI$valid)
```

It gives us 3 metrics: Purity, Fmeasure and the Rand Index (RI). However, we
can obtain this information in another way. Going back to the second element of
`res_with_labels`: `metrics` give us a summary of all metrics:

```{r}
head(res_with_labels$metrics, 3)
```

It gives us the Purity, Fmeasure, RI and Time for every clustering method with
every combination of parameters that has been executed. We can search for
"hierarch_single_euclidean_d2dtaMEId2dtaMHI" and see that it yields the same
results as seen previously:

```{r}
res_with_labels$metrics["hierarch_single_euclidean_d2dtaMEId2dtaMHI", ]
```

But now imagine that we executed `EHyClus` without giving the `true_labels`
parameter but we want to compute the metrics. For that purpose, we can use the
`clustering_validation` function, using the true labels as the first parameter
and the ones generated by the clustering method as the second one:

```{r}
clustering_validation(true_labels, res$cluster$hierarch$hierarch_single_euclidean_d2dtaMEId2dtaMHI$cluster)
```

Note that we are only taking a look at some of the result. Let's see if some
of them have given us better metrics. We are going to sort all the results
based on the RI:

```{r}
head(res_with_labels$metrics[order(res_with_labels$metrics$RI, decreasing = TRUE), ], 5)
```

Indeed, we can see that some methods have given us excellent results,
performing an almost perfect clustering.
